---
title: "Effect of missingness - four items w/ weak correlations"
output: html_document
author: "Pooya Razavi"
date: "last knitted: `r Sys.time()`"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(MASS)
library(tidyverse)
library(GGally)

set.seed(110)

knitr::opts_chunk$set(echo = TRUE)
```

**Goal:** Determine the extent to which missing data can affect the accuracy of the results of a survey. <br>

**Survey parameters:** <br>
_Number of items:_ 4 <br>
_Average item interrelations:_ weak (mean(r) = .2) <br>

**Simulation parameters:** <br>
_Sample sizes:_ 5000, 3000, 1000, 500, and 200 <br>
_Number of iterations per condition:_ 100 <br>
_Missingness range:_ from 10 to 90 percent <br>
_Evaluated outcomes:_ (a) correlation between the items, (b) mean of the four items <br>

# Set up the parameters

```{r}

#correlation matrix
cor_matrix <- rbind(c(1, .25, .3, .1),
                    c(.25, 1, .2, .25),
                    c(.3, .2, 1, .1),
                    c(.1, .25, .1, 1))

cor_matrix

#vector of means
mean_vector <- c(5, 3, 4, 5)

#sample sizes for the simulation
sample_size_options <- c(5000, 3000, 1000, 500, 200)


```


# Simulation

```{r results=FALSE, fig.show='hide'}

for (sample_size in sample_size_options) {

  #simulate the data
  df <- as.data.frame(mvrnorm(n = sample_size, 
                                  mu = mean_vector, 
                                  Sigma = cor_matrix))
  
  
  # Testing the effect of missingness
  
  sim_summary <- data.frame(n = NA,
                            missing = NA,
                            iteration = NA,
                            sum_abs_dev = NA,
                            abs_mean_dif = NA)
  
  missing_prob <- seq(0.1, .9, by = 0.025) #test NA percentage bet. .1 and .9
  
  for (missing in missing_prob) {
  
    for (iteration in 1:100) {
    
    #creating a matrix of NAs
    
      for (i in 5:8) {
          df[, i] <- sample(c("response", NA),
                        replace = TRUE,
                        size = sample_size,
                        prob = c(1 - missing, missing))
        }
    
    #apply NAs to the complete dataset
    df_w_na <- df %>% 
                  dplyr::mutate(V1_na = if_else(V5 == "response", V1, -100),
                                V2_na = if_else(V6 == "response", V2, -100),
                                V3_na = if_else(V7 == "response", V3, -100),
                                V4_na = if_else(V8 == "response", V4, -100)
                                  ) %>% 
                  dplyr::select(contains("_na"))
    
    ##outcome 1: correlation between items
      #correlation matrix for the df with NAs
      cor_matrix_w_na <- cor(df_w_na, use = "pairwise.complete.obs") %>% round(2)
      
      
      #the original correlation matrix
      orig_cor_matrix <- cor(df[, 1:4]) %>% round(2)
      
      #compare the original correlation matrix and the one with NAs
      dif_cor_matrix <- orig_cor_matrix - cor_matrix_w_na
      
      #sum of absolute deviation
      sum_abs_dev <- sum(abs(dif_cor_matrix)) / 2
    
    ##outcome 2: mean of four items  
      #the mean of four items (no NAs)
      complete_mean <- (mean(df[,1]) + mean(df[,2]) + mean(df[,3]) + mean(df[,4]))/4
      #the mean of four items (w/ NAs)
      df_w_na_mean <- (mean(df_w_na[,1], na.rm = T) + mean(df_w_na[,2], na.rm = T) + 
                         mean(df_w_na[,3], na.rm = T) + mean(df_w_na[,4], na.rm = T))/4
      #absolute mean difference bet. complete and sparse datasets
      abs_mean_dif <- abs(complete_mean - df_w_na_mean)
    
    #save the output of this iteration
    sim_summary <- rbind(sim_summary, c(sample_size, missing, iteration, sum_abs_dev, abs_mean_dif))
    
    #print the iteration step (this is important for long simulations)
    print(paste0("simulation step: n = ", sample_size, ", %NA = ", missing, ", iter. = ", iteration))
     
    }
  }

  #save the plots for the 2 outcomes based on the full simulation for the assigned sample size
  summary_plot_cor <- sim_summary[-1,] %>% 
                        dplyr::group_by(missing) %>% 
                        summarise(med_dev = median(sum_abs_dev),
                                  sd = sd(sum_abs_dev)) %>% 
                        ggplot(aes(x = missing, y = med_dev)) +
                              geom_line(color = "darkgray") +
                              geom_point() +
                              ylim(0, 2) +
                              labs(
                                   title = paste0("N = ", sample_size),
                                   x = "Proportion of missing data",
                                   y = "Median sum of absolute deviation") +
                        geom_errorbar(aes(ymin = med_dev - sd/2, ymax = med_dev + sd/2), width=.01,
                           color = "gray") +
                        geom_hline(yintercept = .18, linetype = "dashed", 
                              color = "red", size = .5) +
                        theme_minimal() 
  
  summary_plot_mean <- sim_summary[-1,] %>% 
                        dplyr::group_by(missing) %>% 
                        summarise(med_dev = median(abs_mean_dif),
                                  sd = sd(abs_mean_dif)) %>% 
                        ggplot(aes(x = missing, y = med_dev)) +
                              geom_line(color = "darkgray") +
                              geom_point() +
                              ylim(0, .5) +
                              labs(
                                   title = paste0("N = ", sample_size),
                                   x = "Proportion of missing data",
                                   y = "Median absolute mean difference") +
                        geom_errorbar(aes(ymin = med_dev - sd/2, ymax = med_dev + sd/2), width=.01,
                           color = "gray") +
                        #geom_hline(yintercept = .18, linetype = "dashed", 
                        #      color = "red", size = .5) +
                        theme_minimal() 
  
  print(summary_plot_cor)
  print(summary_plot_mean)
  
  #save the plots as objects to be used outside the loop
  cor_plot_name <- paste0("plot_cor_", sample_size)
  assign(cor_plot_name, summary_plot_cor)
  
  mean_plot_name <- paste0("plot_mean_", sample_size)
  assign(mean_plot_name, summary_plot_mean)

}

```


# Outcome 1: Difference bet. Correlations

For this outcome, I compare the correlation matrix based on the complete data and correlation matrix based on the dataset that has NAs. The larger the differences, the stronger the deviation caused by missing data.

```{r fig.width=6, fig.height=17, warning=FALSE, fig.align='center'}

ggpubr::ggarrange(plot_cor_5000, plot_cor_3000, plot_cor_1000, plot_cor_500, plot_cor_200, 
          ncol = 1, nrow = 5)
```


# Outcome 2: Difference between Means

For this outcome, I compare the mean of 4 items based on the complete data and mean of four items based on the dataset that has NAs. The larger the differences, the stronger the deviation caused by missing data.

```{r fig.width=6, fig.height=17, warning=FALSE, fig.align='center'}

ggpubr::ggarrange(plot_mean_5000, plot_mean_3000, plot_mean_1000, plot_mean_500, plot_mean_200, 
          ncol = 1, nrow = 5)
```

